---
title: "Data Science II: Final Project- RMD and Knitted Files"
output: pdf_document
Name: Jasmin Martinez
Date: 05/12/2025
---
# Read in data:
```{r}
flu = read.csv("severe_flu.csv")
head(flu)
```

# Libraries:
```{r}
library(caret)
library(tidymodels)
library(splines)
library(mgcv)
library(pdp)
library(earth)
library(tidyverse)
library(ggplot2)
library(bayesQR) 
library(dplyr)
library(ISLR)
library(mlbench)
library(randomForest)
library(ranger)
library(gbm)
library(rpart)
library(rpart.plot)
library(party)
library(partykit)
library(pROC)
library(e1071) 
library(kernlab) 
library(ggrepel)
library(corrplot)
library(plotmo)
```
# Factors 
```{r}
flu <- flu %>%
  mutate(
    #gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")),
    #race = factor(race, levels = c(1, 2, 3, 4), labels = c("White", "Asian", "Black", "Hispanic")),
    #smoking = factor(smoking, levels = c(0, 1, 2), labels = c("Never smoked", "Former smoker", "Current smoker")),
    #diabetes = factor(diabetes, levels = c(0, 1), labels = c("No", "Yes")),
    #hypertension = factor(hypertension, levels = c(0, 1), labels = c("No", "Yes")),
    severe_flu = factor(severe_flu, levels = c(0, 1), labels = c("No", "Yes"))
  )

```

# Exploratory analysis: 
```{r}
# observe first couple of rows
head(flu)
str(flu)
summary(flu)

#checking for missing 
colSums(is.na(flu))

# checking for duplicates 
sum(duplicated(flu))
```

## Bar Plot for distribution of severe flu
```{r}
ggplot(flu, aes(x = severe_flu)) + 
  geom_bar(fill = "green", alpha = 0.5) + 
  theme_minimal() + 
  labs(title = "Distribution of Severe Flu Cases",
       x = "Severe Flu",
       y = "Count")

```

## Summarizing continous variables
```{r}
summary(flu[, c("age", "height", "weight", "bmi", "SBP", "LDL")])
```

## Summarizing categorical variables distribution among those with severe fu vs. without severe flu
```{r}
# For gender
flu %>%
  group_by(severe_flu, gender) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(severe_flu) %>%
  mutate(proportion = count / sum(count))

# For race
flu %>%
  group_by(severe_flu, race) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(severe_flu) %>%
  mutate(proportion = count / sum(count))

# For smoking
flu %>%
  group_by(severe_flu, smoking) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(severe_flu) %>%
  mutate(proportion = count / sum(count))

# For diabetes
flu %>%
  group_by(severe_flu, diabetes) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(severe_flu) %>%
  mutate(proportion = count / sum(count))

# For hypertension
flu %>%
  group_by(severe_flu, hypertension) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(severe_flu) %>%
  mutate(proportion = count / sum(count))
```

## Assessing correlation among continuous variables
```{r}
cor(flu[, c("age", "height", "weight", "bmi", "SBP", "LDL")])
```

## Asses relationship between severe flu and continous variables
```{r}
ggplot(flu, aes(x = severe_flu, y = age)) +
  geom_boxplot(fill = "lightpink") +
  theme_minimal() +
  labs(x = "Severe Flu", y = "Age")

ggplot(flu, aes(x = severe_flu, y = height)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(x = "Severe Flu", y = "Height")

ggplot(flu, aes(x = severe_flu, y = weight)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(x = "Severe Flu", y = "Weight")

ggplot(flu, aes(x = severe_flu, y = bmi)) +
  geom_boxplot(fill = "orange") +
  theme_minimal() +
  labs(x = "Severe Flu", y = "BMI")

ggplot(flu, aes(x = severe_flu, y = SBP)) +
  geom_boxplot(fill = "mediumorchid1") +
  theme_minimal() +
  labs(x = "Severe Flu", y = "SBP")

ggplot(flu, aes(x = severe_flu, y = LDL)) +
  geom_boxplot(fill = "red3") +
  theme_minimal() +
  labs(x = "Severe Flu", y = "LDL")
```
# Data Paritioning: 
```{r}
datSplit = initial_split(data = flu, prop = 0.8)
flu_train = training(datSplit)
flu_test = testing(datSplit)
head(flu_train)
head(flu_test)
```

# Part 1: Evaluating whether boosting and SVM provide superior predictive performance compared to simpler models. 

## Simple Model 1: GLM
```{r}
set.seed(1)
#build logistic regression w/ training dataset 
model_logit = glm(severe_flu ~ age + gender + race + smoking + height + weight + bmi +
               diabetes + hypertension + SBP + LDL,
             data = flu_train, family = binomial)

summary(model_logit)

# Predict probabilities and eval. accuracy of predicted values 
pred_probs = predict(model_logit, newdata = flu_test, type = "response")
pred_classes = ifelse(pred_probs > 0.5, 1, 0)
actuals = flu_test[["severe_flu"]]  # safer than using $
cm = table(Predicted = pred_classes, Actual = actuals)
accuracy = sum(diag(cm)) / sum(cm)
print(cm)
print(accuracy)
```
GLM has an accuracy of 0.765. Low. 

## Simple Model 2: Ridge
```{r}
x = model.matrix(severe_flu ~ . - id, flu_train)[,-1]
y = flu_train[, "severe_flu"] 

corrplot(cor(x), method = "circle", type = "full")

# fitting Ridge regression using caret
ctrl1 = trainControl(method = "cv", number = 10)
set.seed(1)
ridge.fit = train(severe_flu ~ . - id, 
                   data = flu_train,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 0,
                                          lambda = exp(seq(6, -5, length = 100))),
                   trControl = ctrl1)

plot(ridge.fit, xTrans = log)
ridge.fit$bestTune
coef(ridge.fit$finalModel, ridge.fit$bestTune$lambda)
```

## Simple Model 3: Lasso
```{r}
set.seed(1)
lasso.fit = train(severe_flu ~ . - id,
                  data = flu_train,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = 1,
                                         lambda = exp(seq(6, -5, length = 100))),
                  trControl = ctrl1)

lasso_tune = lasso.fit$bestTune
plot(lasso.fit, xTrans = log)
lasso_coef = coef(lasso.fit$finalModel, lasso.fit$bestTune$lambda)
```
## Simple Model 4: LDA
```{r}
ctrl3 = trainControl(method = "repeatedcv", repeats = 5,
summaryFunction = twoClassSummary,
classProbs = TRUE)

set.seed(22)

model.lda = train(x = flu_train[, c("age", "gender", "race", "smoking", "height", "weight", "bmi","diabetes", "hypertension", "SBP", "LDL")],  
                  y = flu_train$severe_flu,
                  method = "lda",
                  metric = "ROC",
                  trControl = ctrl3)

print(model.lda)

lda.pred2 = predict(model.lda, newdata = flu_test)
```

## Simple Model 6: MARS
```{r}
set.seed(1)
x = model.matrix(severe_flu ~ age + gender + race + smoking + height + weight + bmi + diabetes + hypertension + SBP + LDL, data = flu_train)[, -1] 

y = flu_train$severe_flu
ctrl1 = trainControl(method = "cv", number = 10)
mars_grid = expand.grid(degree = 1:4,
                        nprune = 2:20)
set.seed(2)
mars.fit = train(x, y, method = "earth",
                 tuneGrid = mars_grid, trControl = ctrl1) 
ggplot(mars.fit)
mars_tune = mars.fit$bestTune
mars_coef = coef(mars.fit$finalModel)

x_test = model.matrix(severe_flu ~ age + gender + race + smoking + height + weight + bmi + diabetes + hypertension + SBP + LDL, data = flu_test)[, -1]
y_test = flu_test$severe_flu
predictions = predict(mars.fit, newdata = x_test)
mse = mean((predictions- y_test)^2)
```

## Simple Model 7: PLS
```{r}
length(setdiff(names(flu_train), c("severe_flu", "id")))

set.seed(2)
pls_fit <- train(severe_flu ~ . - id,
          data = flu_train,
          method = "pls",
          tuneGrid = data.frame(ncomp = 1:11),
          trControl = ctrl1,
          preProcess = c("center", "scale"))
          predy2_pls2 <- predict(pls_fit, newdata = flu_test)
          mean((flu_test$severe_flu - predy2_pls2)^2)

ggplot(pls_fit, highlight = TRUE)
```

## Modle 8: Boosting
```{r}
set.seed(1)
bst = gbm(severe_flu ~ . - id,
          data = flu_train,
          distribution = "gaussian",
          n.trees = 5000,
          interaction.depth = 2,
          shrinkage = 0.005, 
          cv.folds =10)

gbm.perf(bst, method = "cv")
```
```{r}
ctrl = trainControl(method = "cv")

gbm.grid = expand.grid(n.trees = c(100,200,500,1000,2000,5000,10000),
                        interaction.depth = 1:4,
                        shrinkage = c(0.005,0.01,0.05),
                        n.minobsinnode = c(10))

set.seed(1)
gbm.fit = train(severe_flu ~ . - id,
                data = flu_train,
                method = "gbm",
                tuneGrid = gbm.grid,
                trControl = ctrl,
                verbose = FALSE
                )

ggplot(gbm.fit, highlight = TRUE)
```
```{r}
summary(gbm.fit$finalModel, las = 2, cBars = 19, cex.names = 0.6)
```

## Model 9; SVM-Linear 
```{r}
set.seed(1)
radial.tune = tune.svm(severe_flu ~ . - id,
                        data = flu_train,
                        kernel = "radial",
                        cost = exp(seq(1, 7, len = 50)),
                        gamma = exp(seq(-10, -2,len = 20)))
                        plot(radial.tune, transform.y = log, transform.x = log,
                        color.palette = terrain.colors)

radial.tune$best.parameters
best.radial = radial.tune$best.model
summary(best.radial)
pred.radial = predict(best.radial, newdata = flu_test)
confusionMatrix(data = pred.radial,
                reference = testing_data$diabetes)
```

## Model 10: SVM-Radial 
```{r}
set.seed(1)
radial.tune = tune.svm(severe_flu ~ . - id,
                data = flu_train,
                kernel = "radial",
                cost = exp(seq(1, 7, len = 50)),
                gamma = exp(seq(-10, -2,len = 20)))
plot(radial.tune, transform.y = log, transform.x = log,
color.palette = terrain.colors)
```
```{r}
radial.tune$best.parameters

best.radial = radial.tune$best.model
summary(best.radial)

pred.radial = predict(best.radial, newdata = flu_test)
confusionMatrix(data = pred.radial,
reference = flu_test$severe_flu)
```

## Cross Validation for comparison: 
```{r}
set.seed(1)

ctrl_cv = trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

flu_train = flu_train %>% select(-id)

model_list = list()
results = data.frame(Model = character(), ROC = numeric(), stringsAsFactors = FALSE)

# Model 1: GLM
model_list$GLM = train(severe_flu ~ ., data = flu_train, method = "glm", family = "binomial",
                       trControl = ctrl_cv, metric = "ROC")
results = rbind(results, data.frame(Model = "GLM", ROC = max(model_list$GLM$results$ROC)))

# Model 2: Ridge
model_list$Ridge = train(severe_flu ~ ., data = flu_train, method = "glmnet",
                         tuneGrid = expand.grid(alpha = 0, lambda = exp(seq(6, -5, length = 100))),
                         trControl = ctrl_cv, metric = "ROC")
results = rbind(results, data.frame(Model = "Ridge", ROC = max(model_list$Ridge$results$ROC)))

# Model 3: Lasso
model_list$Lasso = train(severe_flu ~ ., data = flu_train, method = "glmnet",
                         tuneGrid = expand.grid(alpha = 1, lambda = exp(seq(6, -5, length = 100))),
                         trControl = ctrl_cv, metric = "ROC")
results = rbind(results, data.frame(Model = "Lasso", ROC = max(model_list$Lasso$results$ROC)))

# Model 4: LDA
model_list$LDA = train(severe_flu ~ ., data = flu_train, method = "lda",
                       trControl = ctrl_cv, metric = "ROC")
results = rbind(results, data.frame(Model = "LDA", ROC = max(model_list$LDA$results$ROC)))

# Model 5: GAM — commented out
# flu_train$age <- as.numeric(flu_train$age)
# model_list$GAM = train(severe_flu ~ s(age) + gender + race + smoking + height + weight +
#                        s(bmi) + diabetes + hypertension + s(SBP) + s(LDL),
#                        data = flu_train, method = "gam",
#                        trControl = ctrl_cv, metric = "ROC")
# results = rbind(results, data.frame(Model = "GAM", ROC = max(model_list$GAM$results$ROC)))

# Model 6: MARS
model_list$MARS = train(severe_flu ~ ., data = flu_train, method = "earth",
                        tuneLength = 10, trControl = ctrl_cv, metric = "ROC")
results = rbind(results, data.frame(Model = "MARS", ROC = max(model_list$MARS$results$ROC)))

# Model 7: PLS
model_list$PLS = train(severe_flu ~ ., data = flu_train, method = "pls",
                       tuneLength = 15, preProcess = c("center", "scale"),
                       trControl = ctrl_cv, metric = "ROC")
results = rbind(results, data.frame(Model = "PLS", ROC = max(model_list$PLS$results$ROC)))

# Model 8: Boosting (GBM)
model_list$GBM = train(severe_flu ~ ., data = flu_train, method = "gbm",
                       trControl = ctrl_cv, metric = "ROC", verbose = FALSE)
results = rbind(results, data.frame(Model = "Boosting", ROC = max(model_list$GBM$results$ROC)))

# Model 9: SVM - Linear
model_list$SVM_Linear = train(severe_flu ~ ., data = flu_train, method = "svmLinear",
                              trControl = ctrl_cv, metric = "ROC")
results = rbind(results, data.frame(Model = "SVM-Linear", ROC = max(model_list$SVM_Linear$results$ROC)))

# Model 10: SVM - Radial
model_list$SVM_Radial = train(severe_flu ~ ., data = flu_train, method = "svmRadial",
                               trControl = ctrl_cv, metric = "ROC")
results = rbind(results, data.frame(Model = "SVM-Radial", ROC = max(model_list$SVM_Radial$results$ROC)))

results = results %>%
  arrange(desc(ROC))

print(results)
```

# Part 2: Developing a predictive risk score (i.e., the predicted probability) that quantifies the chance of experiencing severe flu based on individual participant characteristics.
Given the Ridge regression model had the best ROC, that is the model that will be used to build the predictive risk score. 
```{r}
library(glmnet)

x = model.matrix(severe_flu ~ . - id, flu)[,-1]
y = flu_train[, "severe_flu"] 

ctrl1 = trainControl(method = "cv", number = 10)
set.seed(1)

ridge.fit = train(severe_flu ~ . - id, 
                   data = flu,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 0,
                                          lambda = exp(seq(6, -5, length = 100))),
                   trControl = ctrl1)

flu$predicted_risk_prob = predict(ridge.fit, newx = x, s = "lambda.min", type = "prob")[, "Yes"]
```

# Part 3: Identifying key demographic and clinical factors that predict the risk of severe flu and assessing how these factors influence the risk.
```{r}
# Load library
library(glmnet)

best_lambda = ridge.fit$bestTune$lambda

ridge_glmnet = ridge.fit$finalModel

coef_ridge = coef(ridge_glmnet, s = best_lambda)

coef_df = as.data.frame(as.matrix(coef_ridge))
coef_df$variable <- rownames(coef_df)
colnames(coef_df)[1] <- "coefficient"
coef_df = coef_df %>% 
  filter(variable != "(Intercept)") %>% 
  arrange(desc(abs(coefficient)))

head(coef_df, 10)

library(ggplot2)

# Predicted riskby continuous variable
ggplot(flu, aes(x = age, y = predicted_risk_prob)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess") +
  labs(title = "Predicted Risk of Severe Flu by Age",
       x = "Age", y = "Predicted Risk") +
  theme_minimal()

ggplot(flu, aes(x = height, y = predicted_risk_prob)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess") +
  labs(title = "Predicted Risk of Severe Flu by height ",
       x = "Age", y = "Predicted Risk") +
  theme_minimal()

ggplot(flu, aes(x = bmi, y = predicted_risk_prob)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess") +
  labs(title = "Predicted Risk of Severe Flu by BMI ",
       x = "Age", y = "Predicted Risk") +
  theme_minimal()

ggplot(flu, aes(x = SBP, y = predicted_risk_prob)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess") +
  labs(title = "Predicted Risk of Severe Flu by SBP",
       x = "Age", y = "Predicted Risk") +
  theme_minimal()

ggplot(flu, aes(x = LDL, y = predicted_risk_prob)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess") +
  labs(title = "Predicted Risk of Severe Flu by LDL",
       x = "Age", y = "Predicted Risk") +
  theme_minimal()


library(pdp)

# Partial dependence plots (continous)
pdp_age = partial(ridge.fit, pred.var = "age", train = flu, type = "classification", prob = TRUE)

pdp_height = partial(ridge.fit, pred.var = "height", train = flu, type = "classification", prob = TRUE)

pdp_weight = partial(ridge.fit, pred.var = "weight", train = flu, type = "classification", prob = TRUE)

pdp_bmi = partial(ridge.fit, pred.var = "bmi", train = flu, type = "classification", prob = TRUE)

pdp_SBP=partial(ridge.fit, pred.var = "SBP", train = flu, type = "classification", prob = TRUE)

pdp_LDL=partial(ridge.fit, pred.var = "LDL", train = flu, type = "classification", prob = TRUE)

pdp_gender = partial(ridge.fit, pred.var = "gender", train = flu, type = "classification", prob = TRUE)
plotPartial(pdp_gender, main = "Partial Dependence: Gender")

ggplot(flu, aes(x = factor(gender), y = predicted_risk_prob)) +
  geom_boxplot(fill = "#F28E2B") +
  labs(title = "Distribution of Predicted Risk by Gender",
       x = "Gender", y = "Predicted Risk") +
  theme_minimal()

ggplot(flu, aes(x = factor(race), y = predicted_risk_prob)) +
  geom_boxplot(fill = "blue") +
  labs(title = "Distribution of Predicted Risk by Gender",
       x = "Race (1 = White, 2 = Asian, 3 = Black, 4 = Hispanic)", y = "Predicted Risk") +
  theme_minimal()

ggplot(flu, aes(x = factor(smoking), y = predicted_risk_prob)) +
  geom_boxplot(fill = "pink") +
  labs(title = "Distribution of Predicted Risk by Gender",
       x = "Smoking (0 = Never smoked, 1 = Former smoker, 2 = Current smoker)", y = "Predicted Risk") +
  theme_minimal()

ggplot(flu, aes(x = factor(diabetes), y = predicted_risk_prob)) +
  geom_boxplot(fill ="red") +
  labs(title = "Distribution of Predicted Risk by Gender",
       x = "diabetes (0 = No, 1 = Yes)", y = "Predicted Risk") +
  theme_minimal()

ggplot(flu, aes(x = factor(hypertension), y = predicted_risk_prob)) +
  geom_boxplot(fill ="purple") +
  labs(title = "Distribution of Predicted Risk by Gender",
       x = "hypertension (0 = No, 1 = Yes)", y = "Predicted Risk") +
  theme_minimal()

plotPartial(pdp_age, main = "Partial Dependence: Age")
plotPartial(pdp_height, main = "Partial Dependence: Height")
plotPartial(pdp_weight, main = "Partial Dependence: weight")
plotPartial(pdp_bmi, main = "Partial Dependence: BMI")
plotPartial(pdp_SBP, main = "Partial Dependence: SBP")
plotPartial(pdp_LDL, main = "Partial Dependence: LDL")
```

